Constructing a universe for the setoid model

### 2021-01-22 paper.tex
18p

% TODO notes.
%\usepackage[textsize=small]{todonotes}
%\setlength{\marginparwidth}{2cm}
%\newcommand{\txatodo}[1]{\todo[fancyline,color=green!20]{#1}{}}

@@

The setoid model can be framed categorically as a category with families (CwF,
\cite{internal-tt}) with extra structure for the various type and term
formers. The core structure of a CwF can be given as the following signature:
\vspace{-0.2em}
\begin{align*}
  \Con &: \mType \\
  \Ty &: (\GG : \Con) \to \mType \\
  \Sub &: (\GG\ \GD : \Con) \to \mType \\
  \Tm &: (\GG : \Con) \to \Ty\ \GG \to \mType
\end{align*}

In our presentation of the setoid model, contexts are given by setoids, that is,
types together with an equivalence relation. A key point of the model is that the
equivalence relation is valued in $\mProp$ and is thus definitionally proof
irrelevant.
\[
\infer{
  \begin{aligned}
    | \GG | &: \mType \\
    \GG^\sim &: | \GG | \to | \GG | \to \mProp \\
    \textsf{refl}\ \GG &: (\gamma : | \GG |) \to \GG^\sim\,\gamma\,\gamma \\
    \textsf{sym} \ \GG &: \forall\{\gamma_0\ \gamma_1\} \to \GG^\sim\,\gamma_0\,\gamma_1
    \to \GG^\sim\,\gamma_1\,\gamma_0 \\
    \textsf{trans} \ \GG &: \forall\{\gamma_0\ \gamma_1\ \gamma_2\}
    \to \GG^\sim\,\gamma_0\,\gamma_1
    \to \GG^\sim\,\gamma_1\,\gamma_2
    \to \GG^\sim\,\gamma_0\,\gamma_2
  \end{aligned}
}{
  \GG : \Con
}
\]

Types in a context $\GG$ are given by displayed setoids over $\GG$
with a fibration condition given by $\textsf{coe}$, $\textsf{coh}$. In
the following, we sometimes omit implicit quantifications such as the $\forall\{\gamma_0\ \gamma_1\}$ in the type of $\textsf{sym}\,\GG$.
\[
\infer{
  \begin{aligned}
    | A | &: | \GG | \to \mType \\
    A^\sim &: \{\gamma_0\ \gamma_1 : | \GG |\}
    \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
    \textsf{refl*} &: \{\gamma : | \GG | \} (a : | A | \gamma) \to A^\sim\ (\textsf{refl}\ \GG\ \gamma)\ a\ a \\
    \textsf{sym*} &: \forall\{\gamma_0\,\gamma_1\,a_0\,a_1\}\{p : \GG^\sim\ \gamma_0\ \gamma_1\}
    \to A^\sim\ p\ a_0\ a_1 \to A^\sim\ (\textsf{sym}\ \GG\ p)\ a_1\ a_0 \\
    \textsf{trans*} &: A^\sim\ p_0\ a_0\ a_1\to A^\sim\ p_1\ a_1\ a_2\to A^\sim\ (\textsf{trans}\,\GG\,p_0\,p_1)\ a_0\ a_2\\
    \textsf{coe} &: \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
    \textsf{coh} &: (p : \GG^\sim\ \gamma_0\ \gamma_1) (a : | A | \gamma_0)
    \to A^\sim\,p\,a\,(\textsf{coe}\,A\,p\,a) \\
  \end{aligned}
}{
  A : \Ty\ \GG
}
\]

@@

Substitutions are interpreted as functors between the corresponding setoids,
whereas terms of type $A$ in context $\GG$ are sections of the type seen as a
setoid fibration $\GG . A \to \GG$. Note that we only need to include components
for the functorial action on objects and morphisms, since the functor laws
follow from proof-irrelevance in the metatheory, and thus hold definitionally.
\\
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    | \sigma | &: | \GG | \to | \GD | \\
    \sigma^\sim &: \GG^\sim\, \rho_0\, \rho_1 \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
  \end{aligned}
}{
  \sigma : \Sub\ \GG\ \GD
}
\]
\end{minipage}
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    | t | &: (\gamma : | \GG |) \to | A | \, \gamma \\
    t^\sim &: (p : \GG^\sim\ \gamma_0\ \gamma_1) \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
  \end{aligned}
}{
  t : \Tm\ \GG\ A
}
\]
\end{minipage}
\\

@@

We have a universe of propositions $\Prop$ defined as follows:
\[
\infer{
  \Prop : \Ty\ \GG
}{
  \GG : \Con
}
\qquad
\infer{
  \underline{P} : \Ty\ \GG
}{
  P : \Tm\ \GG\ \Prop
}
\qquad
\infer{
  u \equiv v\vphantom{\Prop}
}{
  u : \Tm\ \GG\ \underline{P} && v : \Tm\ \GG\ \underline{P}
}
\]

Equality type constructors for contexts and dependent types internalize the idea
that every context and type comes equipped with a setoid equivalence
relation. Note that $\mProp$ is the universe of the metatheory while $\Prop$ is
the internal one. As in the model, equality for dependent types is indexed over context
equality.
\[
\infer{
  \GG^\sim\ \rho_0\ \rho_1 : \mProp
}{
  \GG : \Con \qquad \subj{\rho_0, \rho_1}{\DD}{\GG}
}
\qquad
\infer{
  A^{\sim}\ \rho_{01}\ a_0\ a_1 : \Tm\ \DD\ \Prop
}{
  \deduce{
    a_0 : \Tm\ \DD\ A[\rho_0] \qquad a_1 : \Tm\ \DD\ A[\rho_1]
  }{
    A : \Ty\ \GG \qquad \rho_{01} : \GG^{\sim}\ \rho_0\ \rho_1
  }
}
\]

We have rules witnessing that these are indeed equivalence relations. We only
recall reflexivity:
\[
\infer{
  \textsf{R}\ \rho : \GG^\sim\ \rho\ \rho
}{
  \rho : \Sub\ \GD\ \GG
}
\qquad
\infer{
  \textsf{R}\ a : \Tm\ \GG\ \underline{A^\sim\ (\textsf{R}\ \rho)\ a\ a}
}{
  A : \Ty\ \GG
  \qquad
  \rho : \Sub\ \GD\ \GG
  \qquad
  a : \Tm\ \GD\ A[\rho]
}
\]

In addition, we also have rules representing the fact that every construction in
SeTT respects setoid equality, so that we can transport along any such equality:
    \[
    \infer{
      \deduce{
        \tmj{\GD}{\coh_A\ p\ a}{\underline{A^\sim\ p\ a\ (\coe_A\ p\ a)}}
      }{
        \tmj{\GD}{\coe_A\ p\ a}{A[\rho_1]}
      }
    }{
      \tyj{\GG}{A} \qquad \subj{\rho_0, \rho_1}{\GD}{\GG} \qquad
      p : \GG^\sim\ \rho_0\ \rho_1\qquad
      \tmj{\GD}{a}{A[\rho_0]}
    }
    \]

@@

We can easily recover the usual Martin-L\"of identity type from setoid equality,
with transport implemented via coercion.
\[
\infer{
  \textsf{Id}_A\ a_0\ a_1 \defeq A^{\sim}\ (\R\ \GG)\ a_0\ a_1 : \Tm\ \GG\ \Prop
}{
  \tyj{\GG}{A} \qquad \tmj{\GG}{a_0, a_1}{A}
}
\]
\[
\infer{
  \textsf{transp}\ P\ p\ t \defeq
    \coe\ P\ (\textsf{R}\ \textsf{id} , p)\ t : \Tm\ \GG\ P[a_1]
}{
  P : \Ty\ (\GG . A)
  \qquad
  p : \Tm\ \GG\ (\Id\ A\ a_0\ a_1)
  \qquad
  t : \Tm\ \GG\ P[a_0]
}
\]

@@

As an example, consider the following definition of a small universe with
bool and $\Pi$. \\
\vspace{-0.2em}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \textsf{data}\ \U\ : \mType \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
\end{align*}
\end{minipage}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \El\ : \U \to \mType \\
  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) \defeq (a : \El\ A) \to \El\ (B\ a) \\
\end{align*}
\end{minipage}

@@

In our case, we can encode $\U$ as an inductive type $\inU$ that
\emph{carves out} all types in $\mType$ that are in the image of $\El$. In other
words, $\inU$ is a predicate that holds for any type that would have been
obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed by
the type of codes, the definition of $\inU$ quite expectedly reflects the
inductive structure of codes.
\vspace{-0.2em}
\begin{alignat*}{10}
  & \textsf{data}\ \inU && : \mType \to \mType_1 \\
  & \quad \textsf{inBool} && : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi} && :  \inU\ A \to ((a : A) \to \inU\ (B\ a)) \to \inU\ ((a : A) \to (B\ a))
\end{alignat*}

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{sec:setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes, in addition to proofs that these are
indeed equivalence relations: \\
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  \textsf{data} \ \setoidU &: \mType_1 \\
  \blank\sim_\setoidU\blank &: \setoidU \to \setoidU \to \mProp_1
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  \textsf{refl}_\setoidU &: (A : \setoidU) \to \equ{A}{A} \\
  \textsf{sym}_\setoidU &: \equ{A}{B} \to \equ{B}{A} \\
  \textsf{trans}_\setoidU &: \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}
\end{minipage}
\\

$\EL$ is given by a family of setoids indexed over the universe, that is, a way
to assign to each code in the universe a carrier set and an equivalence
relation.
\vspace{-0.2em}
\begin{align*}
  \El &: \setoidU \to \mType \\
  \blank\vdash\blank\sim_\El\blank &: \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
\end{align*}

Note that $\eqel{\blank}{\blank}{\blank}$ is indexed over equality on the universe, because
$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
\begin{align*}
  \textsf{refl}_\El &: (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  \textsf{sym}_\El &: \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  \textsf{trans}_\El &: \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  \coeel &: \equ{A}{B} \to \El\ A \to \El\ B \\
  \cohel &: (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}
%
We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\blank\sim_\setoidU\blank$, $\textsf{refl}_\setoidU$, $\El$ and $\blank\vdash\blank\sim_\El\blank$.
The other functions are then recursively defined:
\begin{itemize}
\item $\textsf{refl}_\El$ alone,
\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
\end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data}\ \setoidU : \mType_1 \\
  & \quad \textsf{bool} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B\ h) \defeq \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}
\end{minipage}
\\

@@

We can now give a full definition of the setoid universe, and of $\EL\ A$ for
any $A : \Tm\ \GG\ \Uty$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  | \Uty | &\defeq \lambda\, \gamma . \, \setoidU \\
  \Uty^\sim &\defeq \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  \textsf{refl}\ \Uty &\defeq \reflu \\
  &\ldots \\
  \textsf{coe}\ \Uty &\defeq \lambda\,p\,a . \, a \\
  \textsf{coh}\ \Uty &\defeq \lambda\,p . \, \reflu
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  | \EL\ A | &\defeq \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
  (\EL\ A)^\sim &\defeq \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
  \textsf{refl}\ (\EL\ A) &\defeq \reflel \\
  &\ldots \\
  \textsf{coe}\ (\EL\ A) &\defeq \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
  \textsf{coh}\ (\EL\ A) &\defeq \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
\end{align*}
\end{minipage}
\\

We can show that $\Uty$ is closed under $\Pi$ types and booleans, and satisfies
$\EL\,(\mathsf{pi}\,A\,B) \equiv \Pi\,(\EL\,A)\,(\EL\,B)$ and
$\EL\,\mathsf{bool} = \mathsf{Bool}$. The universe can be closed under more
constructions if more codes are added to $\setoidU$.

This gives a complete definition of a universe of setoids, which is, however,
inductive-recursive. Moreover, the kind of recursion involved in this definition
is particularly complex, and not obviously reducible to well-understood notions
of induction-recursion like the one described in \cite{dybjer-setzer}. In any
case, we would like to avoid extending the metatheory with any form of
induction-recursion in order to keep the metatheory as small and
essential as possible.

@@

As in the universe example in Section~\ref{ir-universes}, we now define
$\setoidU$ as a $\Sigma$ type, and $\El$ as the corresponding first projection.
\vspace{-0.2em}
\begin{align*}
  \setoidU & : \mType_1 & \El &: \setoidU \to \mType \\
  \setoidU & \defeq \Sigma\ (X : \mType)\ (\inU\ X) & \El &\defeq \pi_1
\end{align*}

@@

Unfortunately this definition only manages to capture the idea, but does not
actually typecheck. In fact, $\blank\sim_{\setoidU}\blank$ should be a $\mProp_1$-valued
relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
$\Sigma$ type shown above clearly is not, since it quantifies over a type of
relations, which is not a proposition.
%
One possible solution is actually quite simple, and it just involves truncating
the $\Sigma$ type above to force it to be in $\mProp_1$.
\vspace{-0.2em}
\begin{align*}
  \blank\sim_{\setoidU}\blank &: \setoidU \to \setoidU \to \mProp_1 \\
  (A , a) \sim_{\setoidU} (B , b) &\defeq
   \| \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \|
\end{align*}

>>>>

\setlength{\abovedisplayskip}{2pt}
\setlength{\belowdisplayskip}{1pt}
% TODO notes.
%\usepackage[textsize=small]{todonotes}
%\setlength{\marginparwidth}{2cm}
%\newcommand{\txatodo}[1]{\todo[fancyline,color=green!20]{#1}{}}

@@

The setoid model can be framed categorically as a category with families (CwF,
\cite{internal-tt}) with extra structure for the various type and term
formers. The core structure of a CwF can be given as the following signature:
{\small\vspace{-0.2em}
\begin{align*}
  \Con &: \mType \\
  \Ty &: (\GG : \Con) \to \mType \\
  \Sub &: (\GG\ \GD : \Con) \to \mType \\
  \Tm &: (\GG : \Con) \to \Ty\ \GG \to \mType
\end{align*}}\vspace{-1em}

In our presentation of the setoid model, contexts are given by setoids, that is,
types together with an equivalence relation. A key point of the model is that the
equivalence relation is valued in $\mProp$ and is thus definitionally proof
irrelevant.
{\small\[
\infer{
  \begin{aligned}
    | \GG | &: \mType \\
    \GG^\sim &: | \GG | \to | \GG | \to \mProp \\
    \textsf{refl}\ \GG &: (\gamma : | \GG |) \to \GG^\sim\,\gamma\,\gamma \\
    \textsf{sym} \ \GG &: \forall\{\gamma_0\ \gamma_1\} \to \GG^\sim\,\gamma_0\,\gamma_1
    \to \GG^\sim\,\gamma_1\,\gamma_0 \\
    \textsf{trans} \ \GG &: \forall\{\gamma_0\ \gamma_1\ \gamma_2\}
    \to \GG^\sim\,\gamma_0\,\gamma_1
    \to \GG^\sim\,\gamma_1\,\gamma_2
    \to \GG^\sim\,\gamma_0\,\gamma_2
  \end{aligned}
}{
  \GG : \Con
}
\]}\vspace{-1em}

Types in a context $\GG$ are given by displayed setoids over $\GG$
with a fibration condition given by $\textsf{coe}$, $\textsf{coh}$. In
the following, we sometimes omit implicit quantifications such as the $\forall\{\gamma_0\ \gamma_1\}$ in the type of $\textsf{sym}\,\GG$.
{\small\[
\infer{
  \begin{aligned}
    | A | &: | \GG | \to \mType \\
    A^\sim &: \{\gamma_0\ \gamma_1 : | \GG |\}
    \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
    \textsf{refl*} &: \{\gamma : | \GG | \} (a : | A | \gamma) \to A^\sim\ (\textsf{refl}\ \GG\ \gamma)\ a\ a \\
    \textsf{sym*} &: \forall\{\gamma_0\,\gamma_1\,a_0\,a_1\}\{p : \GG^\sim\ \gamma_0\ \gamma_1\}
    \to A^\sim\ p\ a_0\ a_1 \to A^\sim\ (\textsf{sym}\ \GG\ p)\ a_1\ a_0 \\
    \textsf{trans*} &: A^\sim\ p_0\ a_0\ a_1\to A^\sim\ p_1\ a_1\ a_2\to A^\sim\ (\textsf{trans}\,\GG\,p_0\,p_1)\ a_0\ a_2\\
    \textsf{coe} &: \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
    \textsf{coh} &: (p : \GG^\sim\ \gamma_0\ \gamma_1) (a : | A | \gamma_0)
    \to A^\sim\,p\,a\,(\textsf{coe}\,A\,p\,a) \\
  \end{aligned}
}{
  A : \Ty\ \GG
}
\]}\vspace{-1em}

@@

Substitutions are interpreted as functors between the corresponding setoids,
whereas terms of type $A$ in context $\GG$ are sections of the type seen as a
setoid fibration $\GG . A \to \GG$. Note that we only need to include components
for the functorial action on objects and morphisms, since the functor laws
follow from proof-irrelevance in the metatheory, and thus hold definitionally.
\\
{\small\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    | \sigma | &: | \GG | \to | \GD | \\
    \sigma^\sim &: \GG^\sim\, \rho_0\, \rho_1 \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
  \end{aligned}
}{
  \sigma : \Sub\ \GG\ \GD
}
\]
\end{minipage}
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    | t | &: (\gamma : | \GG |) \to | A | \, \gamma \\
    t^\sim &: (p : \GG^\sim\ \gamma_0\ \gamma_1) \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
  \end{aligned}
}{
  t : \Tm\ \GG\ A
}
\]
\end{minipage}
\\}\vspace{-1em}

@@

We have a universe of propositions $\Prop$ defined as follows:
{\small\[
\infer{
  \Prop : \Ty\ \GG
}{
  \GG : \Con
}
\qquad
\infer{
  \underline{P} : \Ty\ \GG
}{
  P : \Tm\ \GG\ \Prop
}
\qquad
\infer{
  u \equiv v\vphantom{\Prop}
}{
  u : \Tm\ \GG\ \underline{P} && v : \Tm\ \GG\ \underline{P}
}
\]}\vspace{-1em}

Equality type constructors for contexts and dependent types internalize the idea
that every context and type comes equipped with a setoid equivalence
relation. Note that $\mProp$ is the universe of the metatheory while $\Prop$ is
the internal one. As in the model, equality for dependent types is indexed over context
equality.
{\small\[
\infer{
  \GG^\sim\ \rho_0\ \rho_1 : \mProp
}{
  \GG : \Con \qquad \subj{\rho_0, \rho_1}{\DD}{\GG}
}
\qquad
\infer{
  A^{\sim}\ \rho_{01}\ a_0\ a_1 : \Tm\ \DD\ \Prop
}{
  \deduce{
    a_0 : \Tm\ \DD\ A[\rho_0] \qquad a_1 : \Tm\ \DD\ A[\rho_1]
  }{
    A : \Ty\ \GG \qquad \rho_{01} : \GG^{\sim}\ \rho_0\ \rho_1
  }
}
\]}\vspace{-1em}

We have rules witnessing that these are indeed equivalence relations. We only
recall reflexivity:
{\small\[
\infer{
  \textsf{R}\ \rho : \GG^\sim\ \rho\ \rho
}{
  \rho : \Sub\ \GD\ \GG
}
\qquad
\infer{
  \textsf{R}\ a : \Tm\ \GG\ \underline{A^\sim\ (\textsf{R}\ \rho)\ a\ a}
}{
  A : \Ty\ \GG
  \qquad
  \rho : \Sub\ \GD\ \GG
  \qquad
  a : \Tm\ \GD\ A[\rho]
}
\]}\vspace{-1em}

In addition, we also have rules representing the fact that every construction in
SeTT respects setoid equality, so that we can transport along any such
equality:
{\small
    \[
    \infer{
      \deduce{
        \tmj{\GD}{\coh_A\ p\ a}{\underline{A^\sim\ p\ a\ (\coe_A\ p\ a)}}
      }{
        \tmj{\GD}{\coe_A\ p\ a}{A[\rho_1]}
      }
    }{
      \tyj{\GG}{A} \qquad \subj{\rho_0, \rho_1}{\GD}{\GG} \qquad
      p : \GG^\sim\ \rho_0\ \rho_1\qquad
      \tmj{\GD}{a}{A[\rho_0]}
    }
    \]}\vspace{-1em}

@@

We can easily recover the usual Martin-L\"of identity type from setoid equality,
with transport implemented via coercion.
{\small\[
\infer{
  \textsf{Id}_A\ a_0\ a_1 \defeq A^{\sim}\ (\R\ \GG)\ a_0\ a_1 : \Tm\ \GG\ \Prop
}{
  \tyj{\GG}{A} \qquad \tmj{\GG}{a_0, a_1}{A}
}
\]
\[
\infer{
  \textsf{transp}\ P\ p\ t \defeq
    \coe\ P\ (\textsf{R}\ \textsf{id} , p)\ t : \Tm\ \GG\ P[a_1]
}{
  P : \Ty\ (\GG . A)
  \qquad
  p : \Tm\ \GG\ (\Id\ A\ a_0\ a_1)
  \qquad
  t : \Tm\ \GG\ P[a_0]
}
\]}\vspace{-1em}

@@

As an example, consider the following definition of a small universe with
bool and $\Pi$.

{\small\vspace{-0.4em}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \textsf{data}\ \U\ : \mType \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
\end{align*}
\end{minipage}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \El\ : \U \to \mType \\
  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) \defeq (a : \El\ A) \to \El\ (B\ a) \\
\end{align*}
\end{minipage}}\vspace{-1em}

@@

In our case, we can encode $\U$ as an inductive type $\inU$ that
\emph{carves out} all types in $\mType$ that are in the image of $\El$. In other
words, $\inU$ is a predicate that holds for any type that would have been
obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed by
the type of codes, the definition of $\inU$ quite expectedly reflects the
inductive structure of codes.
{\small\vspace{-0.4em}
\begin{alignat*}{10}
  & \textsf{data}\ \inU && : \mType \to \mType_1 \\
  & \quad \textsf{inBool} && : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi} && :  \inU\ A \to ((a : A) \to \inU\ (B\ a)) \to \inU\ ((a : A) \to (B\ a))
\end{alignat*}}\vspace{-2em}

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{sec:setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes, in addition to proofs that these are
indeed equivalence relations: \\

{\small\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  \textsf{data} \ \setoidU &: \mType_1 \\
  \blank\sim_\setoidU\blank &: \setoidU \to \setoidU \to \mProp_1
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  \textsf{refl}_\setoidU &: (A : \setoidU) \to \equ{A}{A} \\
  \textsf{sym}_\setoidU &: \equ{A}{B} \to \equ{B}{A} \\
  \textsf{trans}_\setoidU &: \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}
\end{minipage}}\vspace{0.4em}

$\EL$ is given by a family of setoids indexed over the universe, that is, a way
to assign to each code in the universe a carrier set and an equivalence
relation.
{\small\vspace{-0.4em}
\begin{align*}
  \El &: \setoidU \to \mType \\
  \blank\vdash\blank\sim_\El\blank &: \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
\end{align*}}\vspace{-1.4em}

Note that $\eqel{\blank}{\blank}{\blank}$ is indexed over equality on the universe, because
$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
{\vspace{-0.4em}\small\begin{align*}
  \textsf{refl}_\El &: (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  \textsf{sym}_\El &: \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  \textsf{trans}_\El &: \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  \coeel &: \equ{A}{B} \to \El\ A \to \El\ B \\
  \cohel &: (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}}\vspace{-1.6em}

We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\blank\sim_\setoidU\blank$, $\textsf{refl}_\setoidU$, $\El$ and $\blank\vdash\blank\sim_\El\blank$.
The other functions are then recursively defined: $\textsf{refl}_\El$ alone,
$\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
$\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
% The other functions are then recursively defined:
% \begin{itemize}
% \item $\textsf{refl}_\El$ alone,
% \item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
% \item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
% \end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$:

{\small\vspace{-1em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data}\ \setoidU : \mType_1 \\
  & \quad \textsf{bool} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B\ h) \defeq \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}
\end{minipage}}\vspace{1em}

@@

We can now give a full definition of the setoid universe, and of $\EL\ A$ for
any $A : \Tm\ \GG\ \Uty$: \\
{\small\vspace{-0.6em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  | \Uty | &\defeq \lambda\, \gamma . \, \setoidU \\
  \Uty^\sim &\defeq \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  \textsf{refl}\ \Uty &\defeq \reflu \\
  &\ldots \\
  \textsf{coe}\ \Uty &\defeq \lambda\,p\,a . \, a \\
  \textsf{coh}\ \Uty &\defeq \lambda\,p . \, \reflu
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  | \EL\ A | &\defeq \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
  (\EL\ A)^\sim &\defeq \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
  \textsf{refl}\ (\EL\ A) &\defeq \reflel \\
  &\ldots \\
  \textsf{coe}\ (\EL\ A) &\defeq \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
  \textsf{coh}\ (\EL\ A) &\defeq \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
\end{align*}
\end{minipage}}
\\

We can show that $\Uty$ is closed under $\Pi$ types and booleans, and satisfies
$\EL\,(\mathsf{pi}\,A\,B) \equiv \Pi\,(\EL\,A)\,(\EL\,B)$ and
$\EL\,\mathsf{bool} = \mathsf{Bool}$. The universe can be closed under more
constructions if more codes are added to $\setoidU$.
%
This gives a complete definition of a universe of setoids, which is, however,
inductive-recursive. Moreover, the kind of recursion involved in this definition
is particularly complex, and not obviously reducible to well-understood notions
of induction-recursion like the one described in \cite{dybjer-setzer}. In any
case, we would like to avoid extending the metatheory with any form of
induction-recursion in order to keep the metatheory as small and
essential as possible.

@@

As in the universe example in Section~\ref{ir-universes}, we now define
$\setoidU$ as a $\Sigma$ type, and $\El$ as the corresponding first projection.
{\small\vspace{-0.2em}
\begin{align*}
  \setoidU & : \mType_1 & \El &: \setoidU \to \mType \\
  \setoidU & \defeq \Sigma\ (X : \mType)\ (\inU\ X) & \El &\defeq \pi_1
\end{align*}}\vspace{-1em}

@@

Unfortunately this definition only manages to capture the idea, but does not
actually typecheck. In fact, $\blank\sim_{\setoidU}\blank$ should be a $\mProp_1$-valued
relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
$\Sigma$ type shown above clearly is not, since it quantifies over a type of
relations, which is not a proposition.
%
One possible solution is actually quite simple, and it just involves truncating
the $\Sigma$ type above to force it to be in $\mProp_1$.
{\small\vspace{-0.2em}
\begin{align*}
  \blank\sim_{\setoidU}\blank &: \setoidU \to \setoidU \to \mProp_1 \\
  (A , a) \sim_{\setoidU} (B , b) &\defeq
   \| \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \|
\end{align*}}\vspace{-1em}
